---
title: "Getting Started with the Communities Package"
author: "Fabio Morea"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
vignette: "%\\VignetteIndexEntry{Getting Started with the Communities Package} %\\VignetteEncoding{UTF-8}
  %\\VignetteEngine{knitr::rmarkdown}\n"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

# Load required packages
library(communities)
library(igraph)
library(dplyr)
library(ggplot2)
```

## Introduction

The `communities` package addresses fundamental reproducibility issues
in network community detection by providing a comprehensive framework
for solution space exploration. Rather than accepting a single algorithm
run, this package systematically explores the complete solution space
through controlled randomization and Bayesian modeling.

### The Problem with Traditional Community Detection

Traditional community detection workflows suffer from three critical
issues:

1.  **Solution multiplicity**: Many networks admit multiple valid
    community structures
2.  **Input ordering bias**: Results depend on arbitrary node/edge
    processing order
3.  **Outlier handling**: Nodes that don't clearly belong to any
    community are poorly managed

This package provides principled solutions to all three problems.

## Core Concepts

### Solution Space Exploration

Instead of running a community detection algorithm once, we run it
multiple times with systematic vertex permutation. This reveals the full
solution space and quantifies the stability of different partitions.

### Bayesian Framework

Solution probabilities are modeled using a Dirichlet-Multinomial
framework with Beta credible intervals, providing principled uncertainty
quantification and convergence assessment.

### Solution Space Taxonomy

Results are classified into five interpretable categories: - **Single**:
One stable solution - **Dominant**: One solution clearly separated from
alternatives - **Multiple**: Several solutions with comparable
probabilities - **Sparse**: Many low-probability solutions (often
indicating instability) - **Empty**: No valid community structure
detected

## Basic Workflow

### Step 1: Create or Load a Network

Let's start with a simple example using a ring of cliques network:

```{r create_network}
# Create a ring of cliques with known community structure
set.seed(42)
graph <- make_ring_of_cliques(n_cliques = 4, 
                              clique_size = 6, 
                              variant = "RC+C")

# Basic network properties
cat("Network properties:\n")
cat("Vertices:", vcount(graph), "\n")
cat("Edges:", ecount(graph), "\n")
cat("Ground truth communities:", max(V(graph)$gt_community), "\n")

# Visualize the network
plot(graph, 
     vertex.color = V(graph)$gt_community,
     vertex.size = 15,
     vertex.label = NA,
     main = "Ring of Cliques Network")
```

### Step 2: Explore the Solution Space

Now we systematically explore the solution space using the Infomap
algorithm:

```{r explore_solution_space}
# Explore solution space with systematic vertex permutation
solution_space_result <- solutions_space_DM(
  graph = graph,
  n_trials = 200,            # Number of algorithm trials
  method = "LV",             # algorithm
  precision_threshold = 0.1, # Convergence threshold
  random_seed = 123         # For reproducibility
)

# View the discovered solutions
knitr::kable(solution_space_result$probabilities, 
             digits = 3,
             caption = "Solution Space Results")
```

### Step 3: Classify the Solution Space

Determine what type of solution space we discovered:

```{r classify_solution_space}
# Classify the solution space type
space_type <- communities::solution_space_type(solution_space_result)
cat("Solution space type:", space_type, "\n")

# Examine convergence
cat("Convergence details:\n")
cat("- Stopped at trial:", solution_space_result$log$stop_trial, "\n")
cat("- Reason:", solution_space_result$log$stop_reason, "\n")
cat("- Final CI width:", tail(solution_space_result$log$max_ci_width, 1), "\n")
```

### Step 4: Quality Assessment

Assess the quality of discovered solutions:

```{r quality_assessment}
# Comprehensive quality assessment
quality_results <- quality_check(graph, solution_space_result)

# Provide interpretive summary
cat("Quality Assessment Summary:\n")
cat("========================\n")
cat("Total solutions found:", nrow(quality_results), "\n")
cat("Valid solutions:", sum(quality_results$valid, na.rm = TRUE), "\n")
cat("Invalid solutions:", sum(!quality_results$valid, na.rm = TRUE), "\n")

if (any(!quality_results$valid)) {
  cat("\nReasons for invalidity:\n")
  invalid_reasons <- quality_results$reason[!quality_results$valid]
  invalid_reasons <- invalid_reasons[invalid_reasons != ""]
  if (length(invalid_reasons) > 0) {
    for (reason in unique(invalid_reasons)) {
      count <- sum(invalid_reasons == reason)
      cat("-", reason, "(", count, "solutions )\n")
    }
  }
}

# Highlight best solutions
if (any(quality_results$valid)) {
  best_solutions <- quality_results[quality_results$valid, ]
  best_solutions <- best_solutions[order(best_solutions$modularity, decreasing = TRUE), ]
  
  cat("\nTop valid solutions by modularity:\n")
  top_3 <- head(best_solutions, 3)
  for (i in 1:nrow(top_3)) {
    cat(sprintf("Solution %d: Modularity=%.3f, Communities=%d, μ=%.3f\n",
                top_3$solution_id[i], top_3$modularity[i], 
                top_3$k[i], top_3$mu[i]))
  }
}
```

## Advanced Analysis

### Building Consensus Communities

When multiple solutions exist, we can build consensus communities:

```{r consensus_analysis}
# Build co-occurrence matrix
co_occurrence_matrix <- co_occurrence(solution_space_result)

# Find consensus communities
consensus_result <- consensus_communities(
  co_occurrence_matrix = co_occurrence_matrix, 
  co_occurrence_threshold = 0.5,
  group_outliers = FALSE
)

# Summarize consensus results
consensus_summary <- consensus_result %>%
  group_by(consensus_community_label) %>%
  summarise(
    community_size = n(),
    avg_confidence = mean(uncertainty_coefficient, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(community_size))

knitr::kable(consensus_summary,
             digits = 3,
             caption = "Consensus Communities Summary")
```

### Outlier Detection

Identify nodes with uncertain community membership:

```{r outlier_analysis}
# Identify potential outliers based on uncertainty
outlier_candidates <- consensus_result %>%
  filter(uncertainty_coefficient > 0.3) %>%  # High uncertainty threshold
  arrange(desc(uncertainty_coefficient))

if (nrow(outlier_candidates) > 0) {
  knitr::kable(outlier_candidates,
               digits = 3,
               caption = "Potential Outlier Nodes")
} else {
  cat("No outliers detected with uncertainty > 0.3\n")
}
```

## Visualization

### Solution Space Evolution

Track how solution probabilities evolve during exploration:

```{r solution_evolution, fig.width=8, fig.height=5}
# Plot convergence of credible interval widths instead
convergence_data <- data.frame(
  trial = 1:solution_space_result$log$stop_trial,
  ci_width = solution_space_result$log$max_ci_width
)

ggplot(convergence_data, aes(x = trial, y = ci_width)) +
  geom_line(color = "steelblue", size = 1) +
  geom_hline(yintercept = solution_space_result$log$precision_threshold, 
             color = "red", linetype = "dashed", alpha = 0.7) +
  labs(title = "Solution Space Exploration Convergence",
       subtitle = paste("Stopped at trial", solution_space_result$log$stop_trial, 
                        "- Reason:", solution_space_result$log$stop_reason),
       x = "Trial Number", 
       y = "Maximum Confidence Interval Width") +
  theme_minimal()
```

```{r}
communities::plot_sol_space_evolution(solution_space_result)
```

### Solution Comparison

Compare the different solutions found:

```{r solution_comparison, fig.width=10, fig.height=8}
# Plot all solutions found
plot_solutions(
  graph, 
  solution_space_result,
  add_node_labels = FALSE,
  add_prob_labels = TRUE,
  add_title = TRUE
)
```

## Comparing Different Algorithms

Let's compare how different algorithms behave on the same network:

```{r algorithm_comparison}
# Compare multiple algorithms
algorithms <- c("IM", "LV", "LD", "WT")
algorithm_results <- list()

for (alg in algorithms) {
  cat("Running", alg, "algorithm...\n")
  algorithm_results[[alg]] <- solutions_space_DM(
    graph = graph,
    n_trials = 30,  # Fewer trials for comparison
    method = alg,
    precision_threshold = 0.15,
    verbose = FALSE,
    random_seed = 123
  )
}

# Compare solution space types
comparison_summary <- data.frame(
  Algorithm = algorithms,
  Solution_Space_Type = sapply(algorithm_results, solution_space_type),
  N_Solutions = sapply(algorithm_results, function(x) ncol(x$partitions)),
  N_Trials = sapply(algorithm_results, function(x) x$log$stop_trial),
  Converged = sapply(algorithm_results, function(x) x$log$stop_reason == "precision_achieved")
)

knitr::kable(comparison_summary,
             caption = "Algorithm Comparison Results")
```

## Real-World Example: Karate Club Network

Let's apply the framework to a classic network dataset:

```{r karate_club_example}
# Load the classic karate club network
karate_graph <- make_graph("Zachary")

# Add edge weights (all equal for this example)
E(karate_graph)$weight <- 1

# Explore solution space
karate_solution_space <- solutions_space_DM(
  graph = karate_graph,
  n_trials = 100,
  method = "LV",  # Louvain algorithm
  precision_threshold = 0.1,
  verbose = FALSE,
  random_seed = 456
)

# Results summary
cat("Karate Club Network Analysis:\n")
cat("Solution space type:", solution_space_type(karate_solution_space), "\n")
cat("Number of solutions found:", ncol(karate_solution_space$partitions), "\n")

# Show solution probabilities
karate_solution_space$probabilities

communities::plot_solutions(karate_graph,karate_solution_space)
```

```{r}
communities::plot_sol_space_evolution(karate_solution_space)
```

 

## Best Practices

### 1. Parameter Selection

**Number of trials**: Start with 50-100 trials. Increase if you get
"Sparse" solution spaces.

**Precision threshold**: Use 0.05 for exploratory analysis, 0.02 for
publication-quality results.

**Algorithm choice**: - **Infomap (IM)**: Good general-purpose
algorithm - **Louvain (LV)**: Fast, good for large networks - **Leiden
(LD)**: Improved version of Louvain - **Walktrap (WT)**: Based on random
walks

### 2. Interpreting Results

**Single solution space**: Safe to report the unique solution.

**Dominant solution space**: Report the dominant solution but
acknowledge alternatives exist.

**Multiple solution space**: Consider consensus communities or report
the uncertainty.

**Sparse solution space**: May indicate lack of clear community
structure.

### 3. Quality Assessment

Always check solution quality:

-   Modularity should be positive (preferably \> 0.3)

-   Mixing parameter should be \< 0.5

-   Communities should be internally connected

### 4. Reproducibility

Always set a random seed for reproducible results:

```{r reproducibility_example}
# Example of reproducible analysis
set.seed(789)
result1 <- solutions_space_DM(graph, n_trials = 20, method = "IM", verbose = FALSE)

set.seed(789)  # Same seed
result2 <- solutions_space_DM(graph, n_trials = 20, method = "IM", verbose = FALSE)

# Results should be identical
identical(result1$probabilities, result2$probabilities)
```

## Common Issues and Solutions

### Issue 1: "Sparse" Solution Spaces

**Problem**: Algorithm finds many different solutions with low
probabilities.

**Solutions**: - Increase number of trials - Try a different algorithm -
Check if the network actually has clear community structure - Consider
that the network may be inherently ambiguous

### Issue 2: All Solutions Invalid

**Problem**: Quality assessment marks all solutions as invalid.

**Solutions**: - Relax quality thresholds (e.g., increase
`mu_threshold`) - Check if the network has meaningful community
structure - Examine the mixing parameter - values \> 0.5 suggest weak
communities

## Summary

The `communities` package provides a principled approach to community
detection that addresses fundamental reproducibility issues. Key
benefits:

1.  **Systematic exploration**: Reveals the complete solution space
    rather than a single result
2.  **Uncertainty quantification**: Provides confidence measures for
    community assignments
3.  **Bias mitigation**: Addresses input ordering bias through
    controlled randomization
4.  **Quality assessment**: Comprehensive evaluation of solution
    validity
5.  **Consensus building**: Methods to synthesize information across
    multiple solutions

This framework is particularly valuable for:

-   Research requiring reproducible results

-   Networks with ambiguous community structure

-   Comparative studies of community detection algorithms

-   Applications where uncertainty quantification is important

## Further Reading

For more details on the theoretical foundation, see:

Morea, F. and De Stefano, D. (2024). "A Comprehensive Framework for
Solution Space Exploration in Community Detection." *Under Review*.
